---
author: "Issues to consider : Sanger Dataset"
output: html_document
---

```{r knitr_options, include=FALSE}
library(knitr)
library(dplyr)
library(tidyr)
opts_chunk$set(echo = FALSE, fig.width = 7,fig.height = 8)
```

1. simulation results : fdr control in all true nulls case.
2. examples where it doesnt controlled
3. logistf vs MH comparison
4. estimating the FDR using EUMODIC high throughput phenotyping data.
    * compare approaches
    * resampling from the control to select a threshold ,considers data structure (litter or batch effects). would control for any issues with the statistical method implemented.
    * from the local dataset level to control across the globally analysis.
    * advantage that it is compatible across both continuous and categorical output 
    * leads us to being able to control the FPR over __all the variables for a gene__ - most common IMPC user question.
  
(1) fdr control

```
                        Outcome  orig mid_p
1  Number.Of.Thoracic.Vertebrae 0.015 0.016
2    Number.Of.Lumbar.Vertebrae 0.033 0.034
3    Number.Of.Pelvic.Vertebrae 0.011 0.011
4    Number.Of.Caudal.Vertebrae 0.024 0.024
5        Transitional.Vertebrae 0.007 0.045
6            Shape.Of.Vertebrae 0.007 0.047
7           Fusion.Of.Vertebrae 0.015 0.057
8                       Maxilla 0.003 0.061**
9                Zygomatic.Bone 0.000 0.000
10 Number.Of.Cervical.Vertebrae 0.025 0.026
11                  Skull.Shape 0.009 0.026
12         Number.Of.Ribs.Right 0.002 0.010
13          Number.Of.Ribs.Left 0.002 0.004
14             Shape.Of.Ribcage 0.008 0.013
15                Shape.Of.Ribs 0.019 0.022
16                  Rib.Fusions 0.000 0.000
17                     Clavicle 0.006 0.074**
18                      Scapula 0.000 0.000
19                      Humerus 0.001 0.001
20                       Radius 0.008 0.008
21                         Ulna 0.005 0.005
22                       Pelvis 0.009 0.016
23                        Femur 0.001 0.054
24                        Tibia 0.000 0.000
25                       Fibula 0.000 0.000
26                       Joints 0.038 0.041
27               Shape.Of.Spine 0.010 0.076**
28                        Teeth 0.033 0.046
29                     Mandible 0.000 0.000
30             Number.Of.Digits 0.004 0.060
31              Digit.Integrity 0.018 0.021
32                 Syndactylism 0.000 0.000
33             Polysyndactylism 0.005 0.005
34              Brachydactylism 0.028 0.028
35                     Kyphosis 0.019 0.019
36                     Lordosis 0.000 0.000
37                    Scoliosis 0.007 0.064**
38             Fusion.Processes 0.006 0.050
39             Caudal.Processes 0.000 0.000
40           Cervical.Processes 0.000 0.000
41             Lumbar.Processes 0.000 0.000
42             Sacral.Processes 0.000 0.000
43           Thoracic.Processes 0.024 0.029
```

(2) examples where it doesnt controlled

Typical tables where fdr is controled

```
> tbl_222_sorted %>% slice(1:3) %>% rowwise() %>% do(print(get_222_ftable(.)))
         Gender Female     Male    
         Group      KO  WT   KO  WT
Deformed                           
Y                    0   2    0   0
N                    7 927    7 922
         Gender Female     Male    

         Group      KO  WT   KO  WT
Deformed                           
Y                    0   2    0   0
N                    7 927    7 922
         Gender Female     Male    

         Group      KO  WT   KO  WT
Deformed                           
Y                    0   2    0   0
N                    7 927    6 922
```
Typical tables where fdr is __not__ controled

```
tbl_222_sorted %>% filter(Outcome=="Clavicle") %>% slice(1:3) %>% rowwise() %>%  do(print(get_222_ftable(.)))

         Gender Female     Male    
         Group      KO  WT   KO  WT
Deformed                           
Y                    0   0    0   3
N                    7 929    7 919

         Gender Female     Male    
         Group      KO  WT   KO  WT
Deformed                           
Y                    0   0    0   3
N                    7 929    7 919

         Gender Female     Male    
         Group      KO  WT   KO  WT
Deformed                           
Y                    0   0    0   3
N                    7 929    6 919

tbl_222_sorted %>% filter(Outcome=="Shape.Of.Spine") %>% slice(1:3) %>% rowwise() %>%  do(print(get_222_ftable(.)))

         Gender Female     Male    
         Group      KO  WT   KO  WT
Deformed                           
Y                    0  32    1  10
N                    7 897    6 912

         Gender Female     Male    
         Group      KO  WT   KO  WT
Deformed                           
Y                    0  32    0  10
N                    6 897    7 912

         Gender Female     Male    
         Group      KO  WT   KO  WT
Deformed                           
Y                    0  32    1  10
N                    7 897    5 912
```

```{r}
load(file="sim_w_effect_results_2.RData")
load(file="sim_w_effect_results_extras_2.RData")

fdr <- apply(out[,,1,,,drop=F],c(4,5,1),mean)
pwr <- apply(out[,,2,,,drop=F],c(4,5,1),mean)

# key <- tbl_K %>% summarise(unique(fam_size.3)) %>% as.data.frame()
# tt <- cbind(key,fdr[,,1])[]
#data.frame(Outcome=tt$Outcome,orig=round(tt$`1`,3),mid_p=round(tt$`3`,3))

par(mfrow=c(2,2))
tbl_p %>% tbl_df() %>% filter(Outcome %in% c("Maxilla","Clavicle","Shape.Of.Spine","Scoliosis")) %>% group_by(Outcome) %>% do(plot(sort(.$p),pch="."))

# cbind(tbl_K,tbl_pdf) %>% select(Outcome,alpha_star) %>% filter(Outcome %in% c("Maxilla","Clavicle","Shape.Of.Spine","Scoliosis")) %>% group_by(Outcome) %>% do(plot(sort(.$alpha_star),pch="."))
# 
# tbl_222_sorted %>% slice(1:3) %>% rowwise() %>% do(print(get_222_ftable(.)))
# tbl_222_sorted %>% filter(Outcome=="Clavicle") %>% slice(1:3) %>% rowwise() %>%  do(print(get_222_ftable(.)))
# tbl_222_sorted %>% filter(Outcome=="Shape.Of.Spine") %>% slice(1:3) %>% rowwise() %>%  do(print(get_222_ftable(.)))
# 
# 
# 
F_sizes_alpha <- tbl_pa_each_sim %>% summarise(F_size=unique(fam_size.1)) %>% dplyr::select(F_size)
F_sizes_Gilb <- tbl_pa_each_sim %>% summarise(F_size=unique(fam_size.2)) %>% dplyr::select(F_size)
```

```{r}
# FDR:
par(mfrow=c(4,4))
for(i in 1:43)
{
  matplot(colnames(fdr[i,c(1,3,5,7),]),t(fdr[i,c(1,3,5,7),]),type="l",sub=paste0("Outcome " ,i," |F|=",F_sizes_alpha[i,1]),xlab="Odds Ratio",ylab="fdr",ylim = c(0,0.07))
  abline(h=0.05,col="grey")
#   if(i %in% c(16,32,43))
#   {
#     # legend("top",inset=0,legend=c("org","mid","rand","rand1"),lty = 1:4,col = 1:4, horiz = TRUE,lwd=2, cex=.5,)
#     # dev.copy2pdf(file = paste0("fdr~OR_alpha_2",i,".pdf") ,height=11,width=11)
#   }
}

# Power
par(mfrow=c(4,4))
for(i in 1:43)
{
  matplot(colnames(pwr[i,c(1,3,5,7),]),t(pwr[i,c(1,3,5,7),]),type="l",sub=paste0("Outcome " ,i," |F|=",F_sizes_alpha[i,1]),xlab="Odds Ratio",ylab="Avg. Power",ylim = c(0,0.2))
  # if(i %in% c(16,32,43)) dev.copy2pdf(file = paste0("Power~OR_alpha_2",i,".pdf") ,height=11,width=11)
}
          
```

### Notations:

* Female: n..1

|  |KO  |WT  |    |
|--|:--:|:--:|:--:|
|1 |n111|n121|n1.1|
|0 |n211|n221|n2.1|
|  |n.11|n.21|n..1|

* Male: n..2

|  |KO  |WT  |    |
|--|:--:|:--:|:--:|
|1 |n112|n122|n1.2|
|0 |n212|n222|n2.2|
|  |n.12|n.22|n..2|

|  |KO |WT |   |
|--|:-:|:-:|:-:|
|1 |n11|n12|n1.|
|0 |n21|n22|n2.|
|  |n.1|n.2|n..|

