---
output: html_document
---
### simulation snippet

Function `swap()` that gets two arguments: the p value `p` and the pdf value at the at the observed value of test statistic `f_n`. The function computes `p_swap`, calculates the threshold `t_max` and outputs a vector of rejections `R` of the size of `p`. 

The second function print the tables for given row of the data frame.

```{r, eval=FALSE}
swap <-  function(p,f_n,q = 0.05)
{
  p_swap <- 1 - p + f_n 
  S <- p_swap <= p
  t_vec <- p
  t_vec[S] <- p_swap[S]
  t_vec <- unique(sort(t_vec))
  t_max <- -1
  for (ti in length(t_vec):1)
  {
    if ( (1 + sum(p_swap[S]<=t_vec[ti]))/max(sum(p[!S]<=t_vec[ti]),1) <= q )
    {
      t_max <- t_vec[ti]
      break
    }
  }
  R <- rep(F,length(p))
  R[!S] <- p[!S]<=t_max
  return(R)
}

get_222_ftable <- function(row)
{
  z <- array(c(row$n111,row$n211,row$n121,row$n221,row$n112,row$n212,row$n122,row$n222), dim=c(2, 2, 2))
  #addmargins(z) 
  dimnames(z) <- list("Deformed"=c("Y","N"),"Group"=c("KO","WT"),"Gender"=c("Female","Male"))
  return(ftable(z,row.vars = 1,col.vars = c(3,2)))
}

```

#### tables notations:

```
* Female: n..1
 
   |  |KO  |WT  |    |
   |--|:--:|:--:|:--:|
   |1 |n111|n121|n1.1|
   |0 |n211|n221|n2.1|
   |  |n.11|n.21|n..1|
   
* Male: n..2
 
   |  |KO  |WT  |    |
   |--|:--:|:--:|:--:|
   |1 |n112|n122|n1.2|
   |0 |n212|n222|n2.2|
   |  |n.12|n.22|n..2|
```

`tbl_sim` (in .RData file) is a data frame containing 20339 rows = 473 (KO) x 43 (Outcome).

In each row there are:

* Some of the 2x2x2 table margins.
* ORIGINALY observed inner values `n111,n112,n121,n122,n211,n221,n212` and `n222`
 I kept these merely for the margins computation in the function that print the tables for a given row: `get_222_ftable()`.
* the cells `n111_sim`, `n112_sim`, resulted in one simulation run.
* an indicator where are the true alternative hypotheses  (in this simulation run odds ratio = 5 where there true alternatives).
* `pdf_list` - the pdf probabilities.
* `supp_list` - the corresponding test statistic values.
* `alpha_star`
* `p` and `f_n` - the p-value and pdf value at the observed test statistic resulted from the simulation.

Some code:

```{r, eval=FALSE}
load("tbl_sim.RData")

names(tbl_sim)
# first table in first outcome
tbl_sim[1,]
get_222_ftable(tbl_sim[1,])

# first table in outcome "Pelvis"
tbl_sim[tbl_sim$Outcome=="Pelvis",][1,]
get_222_ftable(tbl_sim[tbl_sim$Outcome=="Pelvis",][1,])

# second outcome "Number.Of.Lumbar.Vertebrae"

p <- tbl_sim[tbl_sim$Outcome=="Number.Of.Lumbar.Vertebrae","p"]
f_n <- tbl_sim[tbl_sim$Outcome=="Number.Of.Lumbar.Vertebrae","f_n"]
true_alt <- tbl_sim[tbl_sim$Outcome=="Number.Of.Lumbar.Vertebrae","true_alt"]
m1 <- sum(true_alt)

summary(p)
plot(sort(p))

R_vec <- swap(p,f_n)

(R <-  sum(R_vec))

(V <- sum(R_vec & !true_alt))

V/max(R,1) # FDR

(S <- sum(R_vec & true_alt))

S/m1 # Ave power
```






